---
title: "Field Trial Statistics"
author: "Hannah Holmes and Dr. Sarah Lower"
date: "5/15/2022"
output:
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
    toc_depth: 4
    fig_caption: TRUE
editor_options:
  chunk_output_type: console
---

# Goal:

**To test whether:** 

  + in **field trials**, there were significantly more male *P. corrusca* captured in sticky traps with solvent + pheromone versus traps with solvent alone.
  + in **laboratory bioassays**, male *P. corrusca* spent more time with solvent + pheromone lures than lures with solvent alone.
  + to **generate figures** for the main manuscript.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Ready the Workspace, warning=FALSE, message=FALSE}
#clear the workspace
rm(list = ls())

#load necessary packages
library(ggplot2)
library(dplyr)
library(kableExtra)
library(patchwork)
library(cowplot)
library(knitr)
library(ggpubr)
library(rstatix)
library(tidyr)
library(ggpubr)
library(reshape2)
library(ggpattern)
```

## Field Trials

### Pennsylvania Field Trials Data Analysis

#### Step 1: Load data
```{r PA Import the Data}
#load the PA data
Ecorr_Trial_df <- read.csv("Field.Data_5.14.2022.csv", header = TRUE)

#make a nice table
Ecorr_Trial_df_summary <- Ecorr_Trial_df %>%
  group_by(Treatment) %>%
  summarise(mean_males = mean(N_individuals),
            sd_males = sd(N_individuals))

#view the table           
kbl(Ecorr_Trial_df_summary, digits = c(2), col.names = c("Treatment", "Mean males captured", "SD")) %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")
```

#### Step 2: Mann Whitney U (Wilcoxon) Test
  Because data not normally distributed (solvent-only are all 0), will proceed with non-parametric tests.
```{r wilcox PA, warning=FALSE}
#wilcox test, P = 
res <- wilcox.test(N_individuals ~ Treatment, data = Ecorr_Trial_df,
            alternative = c("less"))
res
```

  + P = 0.003645 - yes, the control (solvent-only) treatment catches significantly fewer individuals.

#### Step 3: Boxplot
  Graph shows N males captured in solvent-only and solvent + pheromone traps, with standard deviation. During this trial, only male specimens were caught in traps. 
```{r boxplot PA, fig.width= 3, fig.height=3, fig.align="center"}
#make sure Treatment is a factor
Ecorr_Trial_df <- tibble(Ecorr_Trial_df)
Ecorr_Trial_df$Treatment <- factor (Ecorr_Trial_df$Treatment, levels = c("Solvent", "Solvent + Pheromone"))

#make the stat test table
#stat.test <- Ecorr_Trial_df %>%
#  wilcox_test(N_individuals ~ Treatment, alternative = c("less")) %>%
#  add_significance()
#stat.test <- stat.test %>% add_xy_position(x = "Treatment")

Ecorr_Trial_df_mod <- Ecorr_Trial_df
Ecorr_Trial_df_mod$Treatment <- gsub("Solvent", "S", Ecorr_Trial_df_mod$Treatment)
Ecorr_Trial_df_mod$Treatment <- gsub("Pheromone", "P", Ecorr_Trial_df_mod$Treatment)

#make barplot
#q <- ggbarplot(Ecorr_Trial_df_mod, 
#               x = "Treatment", 
#               y = "N_individuals", 
#               fill = "Treatment", 
#               palette = c("#ff7f00", "#377eb8"), 
#               add = "mean_sd", 
#               xlab = FALSE, 
#               ylab = "N males", 
#               font.y = c(14, "plain", "black"),
#               legend = "none", 
#               ylim = c(0, 30))
#add stats
#q_w_sig <- q + stat_pvalue_manual(stat.test, 
#                       label = "p.signif",
#                       vjust = -1, 
#                       bracket.nudge.y = 1, 
#                       bracket.size = .7)
#q_w_sig

q_w_sig <- ggplot(Ecorr_Trial_df_mod, aes(x = Treatment, y = N_individuals, fill = Treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  theme_classic() +
  xlab("Treatment") +
  ylab("N males") +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) +
  scale_fill_manual(values = c("#ff7f00", "#377eb8")) +
  ylim(c(-1,30))+
  geom_jitter(alpha = 0.9) +
  geom_signif(comparisons=list(c("S", "S + P")), annotations="**",
              y_position = 25, tip_length = .01, vjust=0.4)
q_w_sig
#ggsave("PA_field_trials_2022.pdf", q_w_sig, device = "pdf", width = 58, height = 65, units = c("mm"), dpi = 1200)
#ggsave("PA_field_trials_2022.png", q_w_sig, device = "png", width = 58, height = 65, units = c("mm"), dpi = 1200)
```

### Vermont Field Trials Data Analysis

#### Step 1: Input VT Data 
```{r VT Import the Data}
#load the PA data
VT_Ecorr_Trial_df <- read.csv("VT_Field.Data_5.20.2022.csv", header = TRUE)

#make a nice table
VT_Ecorr_Trial_df_summary <- VT_Ecorr_Trial_df %>%
  group_by(Treatment) %>%
  summarise(mean_males = mean(N_individuals),
            sd_males = sd(N_individuals))

#view the table           
kbl(VT_Ecorr_Trial_df_summary, digits = c(2), col.names = c("Treatment", "Mean males captured", "SD")) %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")
```

#### Step 2: Mann Whitney U (Wilcoxon) Test
  Because data not normally distributed (solvent-only are all 0), will proceed with non-parametric tests.
```{r wilcox VT, warning = FALSE}
#wilcox test, P = 
res <- wilcox.test(N_individuals ~ Treatment, data = VT_Ecorr_Trial_df,
            alternative = c("less"))
res
```

  + P = 0.003645 - yes, the control (solvent-only) treatment catches significantly fewer individuals.

#### Step 3: Boxplot
  Graph shows N males for control and treatment groups, with standard deviation. During this trial, only male specimens were caught in traps. 
```{r boxplot VT, fig.width= 3, fig.height=3, fig.align="center"}
#make sure Treatment is a factor
VT_Ecorr_Trial_df <- tibble(VT_Ecorr_Trial_df)
VT_Ecorr_Trial_df$Treatment <- factor (VT_Ecorr_Trial_df$Treatment, levels = c("Solvent", "Solvent + Pheromone"))

#make the stat test table
#stat.test <- VT_Ecorr_Trial_df %>%
#  wilcox_test(N_individuals ~ Treatment, alternative = c("less")) %>%
#  add_significance()
#stat.test <- stat.test %>% add_xy_position(x = "Treatment")

#change treatment labels to be more compact
VT_Ecorr_Trial_df_mod <- VT_Ecorr_Trial_df
VT_Ecorr_Trial_df_mod$Treatment <- gsub("Solvent", "S", VT_Ecorr_Trial_df_mod$Treatment)
VT_Ecorr_Trial_df_mod$Treatment <- gsub("Pheromone", "P", VT_Ecorr_Trial_df_mod$Treatment)

#make barplot
#r <- ggbarplot(VT_Ecorr_Trial_df_mod, 
#               x = "Treatment", 
#               y = "N_individuals", 
#               fill = "Treatment", 
#               palette = c("black", "white"), 
#               add = "mean_sd", 
#               xlab = FALSE, 
#               ylab = "N males", 
#               font.y = c(14, "plain", "black"),
#               legend = "none", 
#               ylim = c(0, 30))
#add stats
#r_w_sig <- r + stat_pvalue_manual(stat.test, 
#                       label = "p.signif",
#                       vjust = -1, 
#                       bracket.nudge.y = 1, 
#                       bracket.size = .7)
#r_w_sig

r_w_sig <- ggplot(VT_Ecorr_Trial_df_mod, aes(x = Treatment, 
                           y = N_individuals)) +
  geom_boxplot_pattern(aes(pattern_fill = Treatment),
                       fill = c("#ff7f00", "#377eb8"),
                       color = "black",
                       pattern = 'stripe',
                       pattern_fill = "black") +
  theme_classic() +
  xlab("Treatment") +
  ylab("N males") +
  theme(legend.position = "none", 
        axis.title.x = element_blank()) +
  ylim(c(-1,30)) +
  geom_jitter(alpha = 0.9) +
  geom_signif(comparisons=list(c("S", "S + P")),
              annotations="**",
              y_position = 25, 
              tip_length = .01, 
              vjust=0.4)

r_w_sig
#ggsave("VT_field_trials_2022.pdf", r_w_sig, device = "pdf", width = 58, height = 65, units = c("mm"), dpi = 1200)
#ggsave("VT_field_trials_2022.png", r_w_sig, device = "png", width = 58, height = 65, units = c("mm"), dpi = 1200)
```

## Lab Trials

### Step 1: Read in the data
 All Data (N unique males = 8)
```{r read in data lab, message=FALSE, warning=FALSE}
#read in data
Behavioral_Analysis <- read.csv("Behavioral_Analysis_Sheet_Rep1.csv", header = TRUE)

#get data into usable form
Behavioral_Analysis_raw <- tibble(Filename = Behavioral_Analysis$File_Name, Male = Behavioral_Analysis$Male, Duration = Behavioral_Analysis$Total_time_of_behavior_s, Behavior = Behavioral_Analysis$Behavior)

#Sum behaviors by each male
Behavioral_Analysis_summary <- Behavioral_Analysis_raw %>%
  group_by(Male, Behavior) %>%
  summarize(Cumulative_time_s = sum(Duration))

#Make longer to figure out where there are instances of no contact in control or experimental -> replace with 0s
Behavioral_Analysis_summary_wide <- Behavioral_Analysis_summary %>%
  pivot_wider(names_from = Behavior, values_from = Cumulative_time_s)

#Change NAs to zero (if a behavior didn't happen, it wasn't logged)
Behavioral_Analysis_summary_wide <- Behavioral_Analysis_summary_wide %>%
  mutate_at(vars(c("TOUCHING_E", "TOUCHING_C", "MOUNTING_E", "MOUNTING_C", "COPULATING_E")), ~replace_na(.,0))

#There were no copulations with control -> add a column for that
Behavioral_Analysis_summary_wide$COPULATING_C <- 0

#add in columns for total contact time with experimental and control
Behavioral_Analysis_summary_wide <-
  Behavioral_Analysis_summary_wide %>%
  mutate(Total_contact_time_E = sum(TOUCHING_E, MOUNTING_E, COPULATING_E),
         Total_contact_time_C = sum(TOUCHING_C, MOUNTING_C, COPULATING_C))

#convert to long form
Behavioral_Analysis_summary_long <-
  Behavioral_Analysis_summary_wide %>%
  pivot_longer(cols = c("TOUCHING_E", "TOUCHING_C", "MOUNTING_E", "MOUNTING_C", "COPULATING_E", "COPULATING_C", "Total_contact_time_E", "Total_contact_time_C"), 
               names_to = "Behavior_by_trap_type", 
               values_to = "Time")
  
#Separate Behavior column into trap type and behavior
Behavioral_Analysis_summary_long <- Behavioral_Analysis_summary_long %>%
  mutate(Behavior = case_when(
    startsWith(Behavior_by_trap_type, "TOUCHING") ~ "Touching",
    startsWith(Behavior_by_trap_type, "MOUNTING") ~ "Mounting",
    startsWith(Behavior_by_trap_type, "COPULATING") ~ "Copulating",
    startsWith(Behavior_by_trap_type, "Total") ~ "Contact"
    ),
    Trap_type = case_when(
      endsWith(Behavior_by_trap_type, "E") ~ "Solvent + Pheromone",
      endsWith(Behavior_by_trap_type, "C") ~ "Solvent"
    ))

#sum time by male and treatment (trap_type)
first_rep_data_by_male <- Behavioral_Analysis_summary_long %>%
  group_by(Male, Trap_type, Behavior) %>%
  summarize(Time_spent = sum(Time))

#calculate means and std_dev
first_reps_mean_std <- first_rep_data_by_male %>%
  group_by(Trap_type, Behavior) %>%
  summarise(mean_time = mean(Time_spent)/60, std_dev = sd(Time_spent)/60)

#make variable that combines the two
first_reps_mean_std <- first_reps_mean_std %>%
  mutate(Mean_SD = paste0(round(mean_time,2), " (", round(std_dev, 2), ")"))

#select just desired columns
first_reps_mean_std_wide <- first_reps_mean_std[,c(1,2,5)] %>%
  spread(Behavior, Mean_SD)

#reorder the columns
first_reps_mean_std_wide_ordered <- first_reps_mean_std_wide[,c(1,5,4,3,2)]

#control data by itself
control_data_by_male <- first_rep_data_by_male %>%
  spread(Behavior, Time_spent) %>%
  filter(Trap_type == "Solvent")

control_data_by_male_ordered <- control_data_by_male[,c(1,2,6,5,4,3)]

#experimental data by itself
experimental_data_by_male <- first_rep_data_by_male %>%
  spread(Behavior, Time_spent) %>%
  filter(Trap_type == "Solvent + Pheromone")

experimental_data_by_male_ordered <- experimental_data_by_male[,c(1,2,6,5,4,3)]

#display means + sds as kable
big_table <- kbl(first_reps_mean_std_wide_ordered) %>%
  kable_classic(full_width = F, html_font = "Cambria", position = "center")
big_table <- add_header_above(big_table, c(" ", "Mean time (sd)*" = 4))
big_table <- add_footnote(big_table, c(paste0("in minutes; N males = ", nrow(control_data_by_male))), notation = "symbol")
big_table
```

Note: All means are smaller for Control traps when compared to experimental (pheromone) traps. However, the standard deviations are large.

### Step 2: Statistical Tests

#### Total time in contact
  
##### 1. Are the data normally distributed?
```{r contact normal, message=FALSE, warning=FALSE, fig.align="center", fig.width=3, fig.height=3}
#1 - is data normally distributed?
contact_data <- first_rep_data_by_male %>%
  filter(Behavior == "Contact")

contact_data$Male <- factor(contact_data$Male, levels =  unique(contact_data$Male))

ggplot(contact_data, aes(x = Time_spent/60)) +
  geom_histogram() +
  theme_classic() +
  labs(x = "Time spent (min)", y = "N fireflies")

#nope
shapiro.test(contact_data$Time_spent/60)
```

  + No. Shapiro-Wilk test < 0.05. Try log-transforming.

##### 2. After square-root transformation, are the data normally distributed?
```{r contact sqrt, message=FALSE, warning=FALSE, fig.align="center", fig.width=3, fig.height=3}
#try log-transform
contact_data <- contact_data %>%
  mutate(sqrtCDmin = sqrt(Time_spent/60))

#looks better -> a lot of zeros
ggplot(contact_data, aes(x = sqrtCDmin)) +
  geom_histogram() +
  theme_classic() +
  labs(x = "Square-root Time spent (min)", y = "N fireflies")

#nope
shapiro.test(contact_data$sqrtCDmin)
```

  + No. Shapiro-Wilk test < 0.05, even after square-root transform -> lots of zeros, especially in control. Proceed with non-parametrics.

##### 3. Do males spend more time in contact with lures with pheromone? (nonparametric)
  + Paired wilcox test
```{r contact pair wilcox}
#do non-parametric test
#wilcox.test in R contains a "paired" option. 
#p = 0.003906
wilcox.test(Time_spent ~ Trap_type, data=contact_data, alternative = c("less"), paired = TRUE)
```

  + P = 0.003906 - yes, significantly less time spent in contact with solvent-only lures
  
##### 4. Visualizing the data
  + Just contact
```{r contact visual, fig.width=3, fig.height=3, fig.align='center', warning=FALSE}
#plotting contact
t_w_sig <- ggplot(contact_data, aes(x = Trap_type, y = Time_spent/60, fill = Trap_type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(x = Trap_type, y = Time_spent/60)) +
  ylab("Time (min)") +
  ylim(c(-0.5, 12)) +
  theme_classic() +
  theme(axis.title.x = element_blank(), legend.position = "none") +
  scale_x_discrete(labels = c("S", "S + P")) +
  scale_fill_manual(values = c("#377eb8", "#ff7f00")) +
  geom_signif(comparisons=list(c("Control", "Experimental")), annotations="**",
              y_position = 11, tip_length = .01, vjust=0.4)

t_w_sig

#ggsave("contact.pdf", t_w_sig, device = "pdf", width = 58, height = 65, units = c("mm"), dpi = 1200)
#ggsave("contact.png", t_w_sig, device = "png", width = 58, height = 65, units = c("mm"), dpi = 1200)
```

  + Another way to view the data (looking at pairs)  
```{r lab trial plotting pairs, fig.width=3, fig.height=3, fig.align='center'}
#plot pairs
ggplot(contact_data, aes(x = Trap_type, y = Time_spent/60, color = Male, group = Male)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00")) +
  theme_classic() +
  ylab("Time (min)") +
  theme(axis.title.x=element_blank()) +
  scale_x_discrete(labels = c("S", "S + P"))
```

#### By different behaviors
  + Contact and other data, broken down by behavior type: the big plot
```{r big plot}
# big plot
first_rep_data_by_male_just_types<- first_rep_data_by_male %>%
  filter(Behavior == "Touching" | Behavior == "Mounting" | Behavior == "Copulating")
  
ggplot(first_rep_data_by_male_just_types, aes(x = Trap_type, y= Time_spent/60, fill = Trap_type, color = Trap_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.25) +
  geom_jitter(aes(x = Trap_type, y = Time_spent/60)) +
  facet_grid(~Behavior) +
  theme_classic() +
  scale_x_discrete(labels = c("S", "S + P")) +
  theme(axis.title.x=element_blank(), legend.position = "none") +
  ylab("Time (min)") +
  scale_color_manual(values = c("#E8751E", "#007AAE")) +
  scale_fill_manual(values = c("#E8751E", "#007AAE"))
```

##### 1. Does the difference in time spent between control and pheromone traps differ among behavioral types?
  + Friedman test of differences by behavioral type
```{r differences}
first_rep_data_by_male$Behavior <- factor(first_rep_data_by_male$Behavior, levels =  c("Touching", "Mounting", "Copulating", "Contact"))

first_rep_data_by_male$Male <- factor(first_rep_data_by_male$Male, levels = unique(first_rep_data_by_male$Male))

#generate data table with pairs
contact_data2 <- first_rep_data_by_male %>%
  filter(Behavior == "Contact") %>%
  spread(key = Trap_type, value = Time_spent)

touch_data2 <- first_rep_data_by_male %>%
  filter(Behavior == "Touching") %>%
  spread(key = Trap_type, value = Time_spent)

mount_data2 <- first_rep_data_by_male %>%
  filter(Behavior == "Mounting") %>%
  spread(key = Trap_type, value = Time_spent)

copulate_data2 <- first_rep_data_by_male %>%
  filter(Behavior == "Copulating") %>%
  spread(key = Trap_type, value = Time_spent)

diff_df2 <- tibble(Male = contact_data2$Male, Diff_touch = (touch_data2$`Solvent + Pheromone` - touch_data2$Solvent)/60, Diff_mount = (mount_data2$`Solvent + Pheromone` = mount_data2$Solvent)/60, Diff_copulate = (copulate_data2$`Solvent + Pheromone` - copulate_data2$Solvent)/60, Diff_contact = (contact_data2$`Solvent + Pheromone` - contact_data2$Solvent)/60)

#try just looking at diffs
df_freid <- diff_df2 %>%
  melt(id.vars= c("Male"), measure.vars = c("Diff_touch", "Diff_mount", "Diff_copulate", "Diff_contact")) %>%
  filter(variable == "Diff_touch" | variable == "Diff_mount" | variable == "Diff_copulate")

df_freid$Male <- factor(df_freid$Male, levels = unique(df_freid$Male))
df_freid$variable <- factor(df_freid$variable, levels = c("Diff_touch", "Diff_mount", "Diff_copulate"))

friedman.test(y=df_freid$value, group=df_freid$variable, blocks=df_freid$Male)
#NS - p = 0.1044
```
  + No, there is not a significant difference in the amount of time spent near pheromone vs solvent across behavioral types.

  + Visualization
```{r diff viz}
long_df <- diff_df2 %>%
  melt(id.vars= c("Male"), measure.vars = c("Diff_touch", "Diff_mount", "Diff_copulate", "Diff_contact"))

long_df <- long_df %>%
  filter(variable == "Diff_touch" | variable == "Diff_mount" | variable == "Diff_copulate")

ggplot(long_df, aes(x = variable, y = value, color = Male)) +
  geom_point(alpha = 0.8) +
  theme_classic() +
  scale_color_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00")) +
  theme(legend.position = "none") +
  labs(x = "Behavior", y = "Difference in time spent (min)") +
  geom_hline(yintercept = 0, lty = 5, color = "grey") +
  scale_x_discrete(labels = c("Touching", "Mounting", "Copulating", "Contact")) +
  ylim(-1,11)
```

##### 2. Does the time spent at the pheromone trap significantly differ among behavioral types?
  + Friedman test of time @ pheromone trap by behavioral type
```{r warning=FALSE}
#isolate just behaviors, NOT their summation (contact)
first_rep_data_by_male_subset <- first_rep_data_by_male %>%
  filter(Behavior == "Touching" | Behavior == "Mounting" | Behavior == "Copulating")

#isolate just the experimental results
first_rep_data_by_male_subset_exp <- first_rep_data_by_male_subset %>%
  filter(Trap_type == "Solvent + Pheromone")

#make sure male (block) is a factor
first_rep_data_by_male_subset_exp$Male <- factor(first_rep_data_by_male_subset_exp$Male)

#make sure behavior (explanatory variable) is a factor
first_rep_data_by_male_subset_exp$Behavior <- factor(first_rep_data_by_male_subset_exp$Behavior, levels = c("Touching", "Mounting", "Copulating"))

#friedman test
friedman.test(y=first_rep_data_by_male_subset_exp$Time_spent, group=first_rep_data_by_male_subset_exp$Behavior, blocks=first_rep_data_by_male_subset_exp$Male)

#pairwise wilcox follow-up
pairwise.wilcox.test(first_rep_data_by_male_subset_exp$Time_spent, first_rep_data_by_male_subset_exp$Behavior, p.adj = "BH")
```

  + Visualization

    + First, the overall plot and stats
```{r visualizing behavioral types}
ggplot(first_rep_data_by_male_subset_exp, aes(x = Behavior, y = Time_spent/60)) +
  geom_boxplot() +
  theme_classic() +
  ylab("Time (min)") +
  theme(axis.title.x = element_blank()) +
  geom_signif(comparisons=list(c("Mounting", "Copulating")), annotations="*",
              y_position = 11, tip_length = .01, vjust=0.4) +
  geom_signif(comparisons=list(c("Touching", "Copulating")), annotations="*",
              y_position = 12, tip_length = .01, vjust=0.4)
```
   
  + Then, by pairs
    
```{r friedman pairs}
ggplot(first_rep_data_by_male_subset_exp, aes(x = Behavior, y = Time_spent/60, color = Male)) +
  geom_point(alpha= 0.6) +
  theme_classic() +
  ylab("Time (min)") +
  theme(axis.title.x = element_blank()) +
  scale_color_manual(values = c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#fdbf6f", "#ff7f00"))
```





